\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{parskip}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{tcolorbox}
\usepackage{hyperref}
\usepackage{titlesec}
\usepackage{xcolor}
\usepackage{longtable}
\usepackage{array}
\usepackage{listings}

\hypersetup{
    colorlinks=true,
    linkcolor=blue!60!black,
    urlcolor=blue!60!black,
    citecolor=blue!60!black
}

\titleformat{\section}{\Large\bfseries\color{blue!40!black}}{\thesection}{1em}{}
\titleformat{\subsection}{\large\bfseries}{\thesubsection}{1em}{}

\lstset{
    basicstyle=\ttfamily\small,
    breaklines=true,
    frame=single,
    rulecolor=\color{gray!50},
    backgroundcolor=\color{gray!5}
}

\title{\textbf{Cubist: A Self-Improving Chess Engine\\via Agentic Tournament Selection}}
\author{Point72 Hackathon --- 24-hour build \\ \textit{Build Plan \& Technical Specification}}
\date{}

\begin{document}
\maketitle

\hrule
\vspace{1em}

\tableofcontents
\vspace{1em}
\hrule

\section{Executive Summary}

\textbf{Cubist} is a chess engine that improves itself overnight. Rather than retraining a model, we evolve the \emph{scaffolding} around a fixed LLM through an agentic tournament loop. Each generation, a strategist agent asks five questions about how the current champion could be stronger; five builder agents work in parallel to answer them in code; the resulting engines fight a tournament; the winner becomes the new champion. We expect a measurable Elo improvement across 3 generations completed within a 24-hour build window.

This document is the team's single source of truth: it states the problem, defines the architecture, lays out the tech stack and repository, allocates work across five engineers in parallel, and provides an hour-by-hour timeline.

\section{Problem Statement}

\begin{tcolorbox}[colback=gray!5,colframe=gray!50,arc=2mm]
\textbf{Question:} Can an LLM-based chess engine improve itself in a single day --- without retraining the model --- by letting agents propose improvements, build them in parallel, and tournament-select the winner?
\end{tcolorbox}

Chess engine improvement traditionally follows two paradigms: \textbf{handcrafted iteration} (Stockfish-style, where humans tune evaluation functions over years) and \textbf{self-play reinforcement learning} (AlphaZero-style, requiring massive compute and weight updates). Neither is feasible in 24 hours with an LLM as the underlying ``brain.''

We propose a third path: hold the model fixed and evolve everything around it --- the prompt, the search wrapper, the opening book, the time-management policy --- through an agentic Darwinian loop. The hypothesis is that scaffolding-level improvement, validated through head-to-head tournament play, can produce measurable Elo gains in hours rather than months.

\section{The Approach}

\subsection{The Three Roles}

\begin{enumerate}[leftmargin=*]
    \item \textbf{Strategist agent} --- Looks at the current champion (and prior round results) and asks 5 distinct improvement questions, spanning categories: prompt, search, opening book, evaluation, sampling.
    \item \textbf{Builder agents ($\times 5$, in parallel)} --- Each takes one question and modifies the champion's harness to implement that improvement. Output: a new runnable engine.
    \item \textbf{Referee} --- \texttt{python-chess} runs the tournament. 6 players (5 new + old champion), round-robin, both colors. Winner becomes new champion \emph{only if it beats the old champion head-to-head}.
\end{enumerate}

\subsection{One Generation, End to End}

\begin{tcolorbox}[colback=gray!5,colframe=gray!50,arc=2mm]
\texttt{[Champion]} $\rightarrow$ Strategist proposes 5 questions \\
\hspace*{2em}$\rightarrow$ 5 Builders run in parallel, each producing a candidate engine \\
\hspace*{2em}$\rightarrow$ Validation: each candidate plays one legal game; broken builds drop out \\
\hspace*{2em}$\rightarrow$ Tournament: up to 6 players $\times$ round-robin $\times$ both colors $=$ up to 30 games \\
\hspace*{2em}$\rightarrow$ Top scorer faces Champion head-to-head \\
\hspace*{2em}$\rightarrow$ If it wins, it becomes the new Champion. Else, Champion persists. \\
\hspace*{2em}$\rightarrow$ Strategist is told what won/lost and why
\end{tcolorbox}

\subsection{Why This Structure}

\begin{itemize}[leftmargin=*]
    \item \textbf{Separation of concerns:} the strategist explores ideas; the builders execute. Neither gets bogged down in the other's job.
    \item \textbf{Parallelism by design:} 5 builders, all tournament games, and the dashboard all run concurrently --- non-negotiable for a 24-hour budget.
    \item \textbf{Diversity by construction:} 5 distinct categories prevent the strategist from collapsing onto trivial prompt tweaks.
    \item \textbf{Evidence-based selection:} tournament play, not the strategist's opinion, decides what survives. The anti-regression gate prevents permanent decay.
\end{itemize}

\section{Architecture}

\subsection{System Diagram}

\begin{tcolorbox}[colback=gray!5,colframe=gray!50,arc=2mm]
\begin{verbatim}
        +---------------------+
        |  Strategist Agent   |  (Claude Opus -- big-picture reasoning)
        +----------+----------+
                   |  5 questions
        +----------v----------+
        |   Builder Pool      |  (5x parallel Claude Sonnet -- code-gen)
        +----------+----------+
                   |  5 candidate engines
        +----------v----------+
        |   Validator         |  (1 game vs random; reject crashers)
        +----------+----------+
                   |  N <= 5 valid candidates + champion
        +----------v----------+
        |  Tournament Runner  |  (asyncio; round-robin; both colors)
        +----------+----------+
                   |  scores, PGNs, Elo deltas
        +----------v----------+
        |  Selection Gate     |  (anti-regression: must beat champion)
        +----------+----------+
                   |  new champion (or unchanged)
                   v
        +---------------------+      +-------------------+
        |  SQLite Storage     +<---->+  FastAPI Backend  |
        +---------------------+      +---------+---------+
                                               |  WebSocket
                                     +---------v---------+
                                     |  React Dashboard  |
                                     +-------------------+
\end{verbatim}
\end{tcolorbox}

\subsection{Engine Interface}

Every engine --- baseline, candidate, or champion --- conforms to the same interface so the tournament runner is engine-agnostic.

\begin{lstlisting}[language=Python]
class Engine(Protocol):
    name: str
    generation: int
    lineage: list[str]   # parent engine names

    async def select_move(
        self,
        board: chess.Board,
        time_remaining_ms: int,
    ) -> chess.Move: ...
\end{lstlisting}

A builder agent's job is to produce a Python module exposing an \texttt{Engine}-conforming class. The runner imports it dynamically.

\section{Tech Stack}

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Layer} & \textbf{Choice} & \textbf{Why} \\
\midrule
Language (backend) & Python 3.11+ & \texttt{python-chess} is the canonical lib \\
Package mgmt & \texttt{uv} & Fastest install; zero-config lockfile \\
Chess primitives & \texttt{python-chess} & Move gen, PGN, board, legal-move check \\
LLM provider & Anthropic Claude API & Strong reasoning; tool-use ergonomics \\
Strategist model & \texttt{claude-opus-4-7} & Big-picture, infrequent calls \\
Player/builder model & \texttt{claude-sonnet-4-6} & Fast \& cheap; high call volume \\
Concurrency & \texttt{asyncio} + \texttt{httpx} & Non-blocking LLM calls; parallel games \\
Backend API & FastAPI & WebSockets out of the box \\
Persistence & SQLite (via \texttt{sqlmodel}) & Zero setup; survives restarts \\
Frontend & React + TypeScript + Vite & Fast HMR; rich component ecosystem \\
Board UI & \texttt{react-chessboard} & Drop-in animated board \\
Charts & Recharts & Simple Elo curve \\
Styling & Tailwind CSS & Rapid layout for demo \\
Live updates & WebSocket (FastAPI) & Sub-second push of moves and reasoning \\
Process mgmt & \texttt{honcho} or shell script & Run backend + frontend together \\
\bottomrule
\end{tabular}
\end{center}

\section{Repository Structure}

\begin{lstlisting}
cubist-chess-engine/
+-- README.md
+-- proposal.tex / proposal.pdf       # this document
+-- pyproject.toml                    # uv-managed
+-- .env.example                      # ANTHROPIC_API_KEY=...
+-- Procfile                          # honcho: backend + frontend
|
+-- backend/
|   +-- pyproject.toml
|   +-- cubist/
|   |   +-- __init__.py
|   |   +-- config.py                 # env, model IDs, time controls
|   |   +-- engines/
|   |   |   +-- base.py               # Engine Protocol + BaseLLMEngine
|   |   |   +-- baseline.py           # Generation-0 engine
|   |   |   +-- registry.py           # Load/save engines from disk
|   |   |   +-- generated/            # Builder output lands here
|   |   +-- agents/
|   |   |   +-- strategist.py         # Generates 5 improvement questions
|   |   |   +-- builder.py            # Generates an Engine subclass
|   |   |   +-- prompts/              # Versioned prompt templates
|   |   +-- tournament/
|   |   |   +-- referee.py            # Plays one game between two engines
|   |   |   +-- runner.py             # Schedules round-robin in parallel
|   |   |   +-- elo.py                # Rating updates
|   |   |   +-- selection.py          # Anti-regression gate
|   |   +-- orchestration/
|   |   |   +-- generation.py         # The big loop: 1 generation
|   |   |   +-- run.py                # CLI entrypoint: run N generations
|   |   +-- storage/
|   |   |   +-- models.py             # Engine, Game, Generation tables
|   |   |   +-- db.py                 # SQLite session
|   |   +-- api/
|   |       +-- server.py             # FastAPI app
|   |       +-- websocket.py          # Live event stream
|   |       +-- routes.py             # REST: engines, games, generations
|   +-- tests/
|       +-- test_referee.py
|       +-- test_baseline.py
|       +-- test_validator.py
|
+-- frontend/
|   +-- package.json
|   +-- vite.config.ts
|   +-- src/
|       +-- main.tsx
|       +-- App.tsx
|       +-- api/
|       |   +-- client.ts             # REST + WS client
|       +-- components/
|       |   +-- LiveBoard.tsx         # react-chessboard wrapper
|       |   +-- EloChart.tsx          # Recharts line chart
|       |   +-- Bracket.tsx           # Tournament tree view
|       |   +-- StrategistFeed.tsx    # Streaming reasoning text
|       |   +-- GenerationTimeline.tsx
|       +-- hooks/
|           +-- useEventStream.ts
|
+-- scripts/
    +-- run_generation.py             # one-off: run a single generation
    +-- eval_match.py                 # head-to-head N-game match
    +-- seed_baseline.py              # initialize the database
\end{lstlisting}

\section{Workstream Allocation (5 people in parallel)}

To enable real parallelism, \textbf{Hour 0--1 is interface-freeze}: all five owners agree on the \texttt{Engine} Protocol, the SQLite schema, the WebSocket event shape, and the REST endpoints. After that, each workstream owns a directory and integrates through those contracts.

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{@{}p{0.18\textwidth}p{0.32\textwidth}p{0.42\textwidth}@{}}
\toprule
\textbf{Owner} & \textbf{Workstream} & \textbf{Deliverables} \\
\midrule
\textbf{Person A} & Engine core \& baseline & \texttt{engines/base.py} (Protocol), \texttt{baseline.py} (gen-0 engine: prompt + parse SAN), engine registry, dynamic loading from \texttt{generated/}. Owns the contract everything else builds against. \\
\textbf{Person B} & Tournament \& referee & \texttt{referee.py} (one game with time control), \texttt{runner.py} (parallel round-robin via \texttt{asyncio.gather}), \texttt{elo.py}, \texttt{selection.py} (anti-regression gate). \\
\textbf{Person C} & Agents (strategist + builder) & \texttt{strategist.py} (Opus, structured output of 5 questions), \texttt{builder.py} (Sonnet, generates Python module conforming to \texttt{Engine}), validator that smoke-tests each candidate. \\
\textbf{Person D} & Frontend dashboard & React app, live board, Elo chart, strategist feed, bracket. Mocks against fake WebSocket events on day 1; switches to real backend once Person E ships the API. \\
\textbf{Person E} & Infra, API, orchestration, demo & FastAPI server, WebSocket event bus, SQLite schema, the top-level \texttt{generation.py} loop, deployment script, demo rehearsal. \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Interface Contracts (frozen at Hour 1)}

\textbf{1. Engine Protocol} (shared by A, B, C):
\begin{lstlisting}[language=Python]
class Engine(Protocol):
    name: str
    generation: int
    lineage: list[str]
    async def select_move(
        self, board: chess.Board, time_remaining_ms: int
    ) -> chess.Move: ...
\end{lstlisting}

\textbf{2. SQLite tables} (E owns; A, B, C write):
\begin{itemize}[leftmargin=*]
    \item \texttt{engines(id, name, generation, parent\_name, code\_path, created\_at)}
    \item \texttt{games(id, white\_id, black\_id, pgn, result, generation, created\_at)}
    \item \texttt{generations(id, number, champion\_before, champion\_after, strategist\_questions\_json, started\_at, finished\_at)}
\end{itemize}

\textbf{3. WebSocket events} (E publishes, D consumes):
\begin{itemize}[leftmargin=*]
    \item \texttt{generation.started \{number, champion\}}
    \item \texttt{strategist.question \{index, text, category\}}
    \item \texttt{builder.completed \{question\_index, engine\_name, ok\}}
    \item \texttt{game.move \{game\_id, fen, san, white, black\}}
    \item \texttt{game.finished \{game\_id, result, pgn\}}
    \item \texttt{generation.finished \{number, new\_champion, elo\_delta\}}
\end{itemize}

\section{Hour-by-Hour Timeline (24h)}

\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{longtable}{@{}p{0.08\textwidth}p{0.92\textwidth}@{}}
\toprule
\textbf{Hour} & \textbf{Milestone} \\
\midrule
\endhead
\textbf{0--1} & \textbf{Kickoff \& interface freeze.} All 5 align on Engine Protocol, DB schema, WS events. Repo scaffolded; \texttt{uv sync} works; FastAPI hello-world up; React dev server up. \\
\textbf{1--4} & \textbf{Parallel build, round 1.} A: baseline engine plays a legal game vs random. B: referee plays one game between two engines. C: strategist returns 5 hardcoded questions; builder takes a question and writes a no-op Engine subclass. D: dashboard renders mocked board + chart from fake events. E: SQLite schema live; WebSocket event bus emits stubs. \\
\textbf{4--7} & \textbf{Vertical slice.} End-to-end: baseline plays itself $\rightarrow$ stored in DB $\rightarrow$ rendered on dashboard. B's tournament runner schedules a 2-engine round-robin in parallel. C's builder produces a real working variant (e.g.\ ``add opening book''). \\
\textbf{7--10} & \textbf{First full generation.} Strategist $\rightarrow$ 5 builders $\rightarrow$ validator $\rightarrow$ tournament $\rightarrow$ selection $\rightarrow$ new champion. Persisted to DB and reflected on dashboard. Latency-tune: cut time-per-move so a generation completes in under 90 minutes. \\
\textbf{10--13} & \textbf{Generation 2 + hardening.} Run a second generation autonomously. Add the anti-regression gate. Fix the inevitable bugs in builder code-gen (sandbox the import, catch syntax errors, retry). Add timeouts everywhere. \\
\textbf{13--16} & \textbf{Generation 3.} Should run with little intervention. Person E starts demo rehearsal. Person D polishes UI: Elo curve animation, strategist reasoning streamed character-by-character. \\
\textbf{16--19} & \textbf{Eval \& evidence.} Run the headline match: gen-3 champion vs gen-0 baseline, 10 games. Capture PGNs and a results table for the demo. Bug-fix marathon. \\
\textbf{19--22} & \textbf{Demo polish.} Tighten the narrative: a 3-minute live walkthrough that shows one generation running end-to-end. Pre-record a backup video in case the API is slow on demo day. Write a 1-page handout. \\
\textbf{22--24} & \textbf{Buffer \& sleep.} Final rehearsal, push to GitHub, deploy a static demo URL if possible. Get some sleep before the judging. \\
\bottomrule
\end{longtable}
\end{center}

\section{What Success Looks Like}

\begin{itemize}[leftmargin=*]
    \item \textbf{Primary:} Final champion beats generation-0 baseline $>60\%$ in a 10-game match.
    \item \textbf{Secondary:} Elo trends upward across generations (not flat, not zigzag).
    \item \textbf{Tertiary:} At least one ``surprising'' winning strategy --- something the team didn't anticipate when listing examples.
    \item \textbf{Demo wow-factor:} Judges watch the strategist reason aloud --- e.g.\ \textit{``Last round, the opening-book variant won, so this round let's ask whether extending it to 12 moves helps, and whether the threat-detection prompt would compose with it.''}
\end{itemize}

\section{Risks \& Mitigations}

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{@{}p{0.28\textwidth}p{0.66\textwidth}@{}}
\toprule
\textbf{Risk} & \textbf{Mitigation} \\
\midrule
LLM move latency dominates the budget & Aggressive parallelism (\texttt{asyncio.gather}); short time control (15--30s/move); cap games per pairing at 2; use Sonnet, not Opus, for moves. \\
Variance --- noisy rankings & Both colors per pairing; require head-to-head win vs champion to crown new champion; report results with explicit game counts. \\
Builder produces broken code & Sandboxed import + smoke test (one game vs random); on failure, drop the candidate and continue with N$<$5. Never block the loop. \\
Strategist mode-collapse & Force category coverage in the prompt (prompt / search / book / eval / sampling); provide history of prior questions to encourage novelty. \\
API rate limits & Retry with exponential backoff; pre-flight check at hour 0; have a backup API key. \\
Demo-day flakiness & Pre-record a backup video at hour 21; have a snapshot DB to replay if live runs fail. \\
Integration hell at hour 12 & Hour-1 interface freeze; daily 30-min sync; everyone mocks against the contracts from day 1. \\
\bottomrule
\end{tabular}
\end{center}

\section{Demo Plan}

\textbf{3-minute live walkthrough:}
\begin{enumerate}[leftmargin=*]
    \item \textbf{Hook (20s):} Show the gen-3 champion beating the gen-0 baseline in 4 moves. ``This engine taught itself this opening overnight.''
    \item \textbf{The loop (90s):} Trigger a fresh generation live. Strategist's questions stream onto the screen. Five builder agents go to work. Tournament games play out in parallel on a mini-bracket. New champion crowned.
    \item \textbf{The evidence (45s):} Elo curve across 3 generations. Headline 10-game match result. One ``surprising winning strategy'' anecdote.
    \item \textbf{The pitch (25s):} ``This is a general recipe --- evolve the scaffolding, not the model. It works for chess; it works for any agent task with a clear win condition.''
\end{enumerate}

\textbf{Backup:} a pre-recorded video of a successful generation at hour 21, in case the live API is slow or rate-limited during judging.

\end{document}
